---
title: "Children's language input"
author: "George"
date: "9/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
```

# Goal

Use average hourly input rates from three sources (overheard speech, child-directed speech, and child-read books) to estimate how many tokens a 'standard' child will experience over their early development.

## Overheard and Child-directed speech

Using children's median number of waking hours, we extrapolate from our best estimates of hourly rates for overheard speech (600 words/hr) and child-directed speech (1198 words/hr) to the monthly amounts of speech heard by children.
(Should these rates change for school-aged children?)

```{r waking-hours}
# https://www.sleepfoundation.org/children-and-sleep/how-much-sleep-do-kids-need
waking_hours_per_day = c(7,8,9, # 14-17 hrs of sleep
                           seq(10, 12, length.out=9), # infant: 12-15 hrs of sleep
                           rep(12, 12), # toddler: 11-14 hrs of sleep
                           rep(13, 3*12), # preschool (3-5): 10-13 hrs of sleep
                           rep(14, 8*12)) # school-aged: 9-11 hrs of sleep
```


```{r hourly-rates}
overheard_speech_rate = rep(600, 13*12) # 13 years - increase at school age?
child_directed_speech_rate = rep(1198, 13*12) # how should this change?
# let's estimate monthly rates based on Hasbrauck and Tindal (2006) norms
reading_rate = 60 * c(rep(0, 12*5), # no reading 0-5
                 seq(0, 53, length.out=12), # 53 correct wpm by end of Grade 1
                 seq(53, 89, length.out=12), # 89 by end of Grade 2
                 seq(89, 107, length.out=12), # 107 by end of Grade 3
                 seq(107, 123, length.out=12), # Grade 4
                 seq(123, 139, length.out=12), # Grade 5
                 seq(139, 150, length.out=12), # Grade 6
                 seq(150, 150, length.out=12), # Grade 7
                 seq(150, 151, length.out=12) # Grade 8
                 ) # first 13 years

# 6TH-8TH GRADE (SPRING) (11-14 YEARS OLD): 150 – 204 WPM
# HIGHSCHOOL (14-18 YEARS OLD): 200 – 300 WPM
# COLLEGE (18-23) YEARS OLD	300 – 350 WPM
# ADULTS	220 – 350 WPM
```

## Reading rates

Many sources recommend that school-aged children read 15-20 minutes per day (7.6 hours/month), although most teachers estimate that most of their students read less than this.
For now we will assume the 'standard' child reads the recommended daily amount, at speeds following the Hasbrauck and Tindal (2006) norms (e.g., 53 correct words/min by end of Grade 1; 151 correct words/min by end of Grade 8).
For reference, silent-reading adults average 238 words/min, while adults that read aloud average 183 words/min (Brysbaert, 2019).

```{r, echo=FALSE}
hours_read_per_month = 30.44*.25 # 15 min/day

get_input_df <- function(overheard_speech_rate, child_directed_speech_rate, 
                         reading_rate, hours_read_per_month, waking_hours_per_day) {
  hours_overheard_speech_per_month = overheard_speech_rate * waking_hours_per_month
  hours_child_directed_speech_per_month = child_directed_speech_rate * waking_hours_per_month

  tokens_read_per_month = reading_rate * hours_read_per_month
  waking_hours_per_month = waking_hours_per_day * 30.44
  
  input_long_df <- bind_rows(
    tibble(Source = "Reading", Age = 1:length(tokens_read_per_month), 
           `Hourly Rate` = reading_rate,
           `Monthly Hours` = hours_read_per_month,
           `Monthly Tokens` = tokens_read_per_month,
           `Total Tokens` = cumsum(tokens_read_per_month)),
    tibble(Source = "Overheard Speech", Age = 1:length(tokens_read_per_month),
           `Hourly Rate` = overheard_speech_rate,
           `Monthly Hours` = waking_hours_per_month,
           `Monthly Tokens` = hours_overheard_speech_per_month,
           `Total Tokens` = cumsum(hours_overheard_speech_per_month)),
    tibble(Source = "Child-directed Speech", Age = 1:length(tokens_read_per_month),
           `Hourly Rate` = child_directed_speech_rate,
           `Monthly Hours` = waking_hours_per_month,
           `Monthly Tokens` = hours_child_directed_speech_per_month,
           `Total Tokens` = cumsum(hours_child_directed_speech_per_month))
    )
  return(input_long_df)
}

input_long_df <- get_input_df(overheard_speech_rate, 
                              child_directed_speech_rate, 
                              reading_rate, 
                              hours_read_per_month, 
                              waking_hours_per_day)
```


## Monthly input

For reference, the average novel is ~60,000 words.

```{r monthly-input, echo=F}
input_long_df %>%
  mutate(`Age (years)` = Age / 12) %>%
  ggplot(aes(x=`Age (years)`, y=`Monthly Tokens`, fill=Source)) + 
    geom_area() + #ylim(0, 1.5e6) + 
    theme_classic() # + geom_hline(yintercept=1e6, linetype="dashed")
```

## Cumulative input

How many tokens do children accumulate from different sources by the age of 13?


```{r, echo=F}
input_long_df %>% 
  mutate(`Cumulative Tokens (millions)` = `Total Tokens` / 1e6,
         `Age (years)` = Age / 12) %>%
  ggplot(aes(x=`Age (years)`, y=`Cumulative Tokens (millions)`, fill=Source)) + 
    geom_area() + 
    theme_classic() + # + geom_hline(yintercept=1e6, linetype="dashed")
    theme(legend.position = c(.2, .7))
```


```{r}
age13 = subset(input_long_df, Age==156)
age13[1,]$`Total Tokens` / sum(age13$`Total Tokens`)
```

## Strong Reader

What if a child reads an hour per day, and is an accelerated reader (starts reading at age 5, and reads at norms 1 grade level ahead)?

```{r}
reading_rate_accel = 60 * c(rep(0, 12*4), # no reading 0-4
                 seq(0, 53, length.out=12), # 53 correct wpm by end of kindergarten
                 seq(53, 89, length.out=12), # 89 by end of Grade 1
                 seq(89, 107, length.out=12), # 107 by end of Grade 2
                 seq(107, 123, length.out=12), # Grade 3
                 seq(123, 139, length.out=12), # Grade 4
                 seq(139, 150, length.out=12), # Grade 5
                 seq(150, 150, length.out=12), # Grade 6
                 seq(150, 151, length.out=12), # Grade 7
                 seq(151, 204, length.out=12)
                 ) # first 13 years


# 6TH-8TH GRADE (SPRING) (11-14 YEARS OLD): 150 – 204 WPM
# HIGHSCHOOL (14-18 YEARS OLD): 200 – 300 WPM
# COLLEGE (18-23) YEARS OLD	300 – 350 WPM
# ADULTS	220 – 350 WPM

input_long_df_accel <- get_input_df(overheard_speech_rate, 
                              child_directed_speech_rate, 
                              reading_rate_accel, 
                              30.44, 
                              waking_hours_per_day)

```

```{r, echo=F}
input_long_df_accel %>% 
  mutate(`Cumulative Tokens (millions)` = `Total Tokens` / 1e6,
         `Age (years)` = Age / 12) %>%
  ggplot(aes(x=`Age (years)`, y=`Cumulative Tokens (millions)`, fill=Source)) + 
    geom_area() + 
    theme_classic() + # + geom_hline(yintercept=1e6, linetype="dashed")
    theme(legend.position = c(.2, .7)) + 
  ggtitle("An Avid Reader")
```

## TV Watching

"The average across all 170 stations is 2.38 words per second" (143 words/min, 8568 words/hr). 
Children ages 2 to 5 watched TV on average 32 hours each week (4.6 hrs/day), and children ages 6 to 8 spent 28 hours per week watching television (Nielson Co., 2009).
The average child (aged 8-12) watches 4-6 hours/day, providing 34320 to 51480 daily words that are not overheard/child-directed speech.

We show cumulative tokens for children watching the average amount of television for their age.
We also assume TV time eats into CDS and ODS equally (i.e., for each hour of TV watched, children lose 30 minutes of CDS, and 30 minutes of ODS).
We assume children younger than 3 do not watch any TV.

```{r, echo=F}
tv_hourly_tokens = 2.38 * 60 * 60

hours_tv_per_day = c(rep(0, 2*12),
                       rep(4.6, 3*12),
                       rep(4.0, 2*12),
                       rep(5.0, 6*12))

tokens_tv_per_month = tv_hourly_tokens * hours_tv_per_day * 30.44

# incorporate this into get_input_df function? or make adjust_for_activity() ?
input_long_tv_df <- bind_rows(input_long_df,
    tibble(Source = "TV", Age = 1:length(tokens_tv_per_month), 
           `Hourly Rate` = tv_hourly_tokens,
           `Monthly Hours` = hours_tv_per_day * 30.44,
           `Monthly Tokens` = tokens_tv_per_month,
           `Total Tokens` = cumsum(tokens_tv_per_month)))

input_wide_tv_df <- input_long_tv_df %>% 
  pivot_wider(id_cols=Age, names_from=Source, values_from=`Monthly Hours`) %>%
  mutate(`Overheard Speech` = `Overheard Speech` - TV/2,
         `Child-directed Speech` = `Child-directed Speech` - TV/2)

input_wide_tv_df
```

```{r, echo=F}
input_long_tv_df %>% 
  mutate(`Cumulative Tokens (millions)` = `Total Tokens` / 1e6,
         `Age (years)` = Age / 12) %>%
  ggplot(aes(x=`Age (years)`, y=`Cumulative Tokens (millions)`, fill=Source)) + 
    geom_area() + 
    theme_classic() + # + geom_hline(yintercept=1e6, linetype="dashed")
    theme(legend.position = c(.2, .7))
```

