---
title             : "The ``Standard Model'' of Early Language Learning"
shorttitle        : "Standard Model of Language Learning"

author: 
  - name          : "George Kachergis"
    affiliation   : "1"
    corresponding : no 
    address       : "420 Serra Mall"
    email         : "kachergis@stanford.edu"
  - name          : "Michael C. Frank"
    affiliation   : "1"
    corresponding : yes 
    address       : "420 Serra Mall"
    email         : "mcfrank@stanford.edu"

affiliation:
  - id            : "1"
    institution   : "Stanford University"

authornote: |
  Department of Psychology

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
  
  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
  
  One sentence clearly stating the **general problem** being addressed by  this particular study.
  
  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
  
  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
  
  One or two sentences to put the results into a more **general context**.
  
  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
  
  
keywords          : "keywords"
wordcount         : "X"

bibliography      : ["r-references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

In physics, the "Standard Model" is a widely-accepted theory completed in the 1970s that describes all known elementary particles, along with three of the four known fundamental forces in the universe, explaining a wide variety of empirical phenomena. 
What fundamental building blocks and mechanisms are needed to construct an analogous standard model of early language learning?

Understanding children's language environment has been of central interest to both theoretical and applied and research on early language learning.
In particular, word learning has come to be viewed as central to measuring and influencing language learning.
Important applied area, interest in sociodemographic variation

Word learning is one of the rare areas in developmental work in which the construct of individual differences is also seen as an epistemic construct: 


# The Standard Model

To build a "standard model", we begin by specifying the minimal assumptions made by any model of word learning.
First, assume that words accumulate through repeated exposure at a rate controlled by a variety of factors, including 1) learning rate, 2) frequency of exposure, 3) diversity of exposure, and 4) leveraged learning.

Combine to create a "macro-economics of early language" 
Input distributional patterns that match corpus dynamics (Montag etc)
Identify critical uptake patterns to aim for (CDI quartiles?)

## Extensions
ambiguity and disambiguation
Cross-situational extensions (Smith, Blythe) - most word learning situations are ambiguous and require disambiguation across time (using mechanisms of learning and memory)
Socio-pragmatic extensions - most word learning instances are disambiguated by social and pragmatic cues (and are unambiguous)
Experimental evidence from the "associations vs. hypotheses" debate


# Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.


# Results

# Discussion


\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup
