---
title             : "The ``Standard Model'' of Early Language Learning"
shorttitle        : "Standard Model of Language Learning"

author: 
  - name          : "George Kachergis"
    affiliation   : "1"
    corresponding : no 
    address       : "420 Jane Stanford Way"
    email         : "kachergis@stanford.edu"
  - name          : "Michael C. Frank"
    affiliation   : "1"
    corresponding : yes 
    address       : "420 Jane Stanford Way"
    email         : "mcfrank@stanford.edu"

affiliation:
  - id            : "1"
    institution   : "Stanford University"

authornote: |
  Department of Psychology

abstract: |
  While there remains vigorous debate about the mechanisms by which children learn their first words, there is an emerging consensus about some criticcal factors that predict acquisition.
  For example, although some words are more difficult to learn than others, words that are heard more often are generally learned earlier: word frequency matters.
  Moreover, the overall amount of input heard by a child--whether child-directed or overheard speech--should influence uptake: a child hearing more hourly tokens should learn words earlier.
  Building on previous efforts to model children's early word learning, we present a "standard model" that makes minimal, universally-agreed-upon assumptions about the learning process.
  Using empirically-estimated input rate distributions and per-word frequencies from the CHILDES corpus, we produce simulated trajectories of vocabulary development.
  Finally, we discuss implications of the standard model for designing and simulating interventions, as well as outstanding parameter identifiability issues and directions for further model develpment and testing.
  
keywords          : "keywords"
wordcount         : "X"

bibliography      : ["references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

In physics, the "Standard Model" is a widely-accepted theory completed in the 1970s that describes all known elementary particles, along with three of the four known fundamental forces in the universe, explaining a wide variety of empirical phenomena. 
What fundamental building blocks and mechanisms are needed to construct an analogous standard model of early language learning?

Understanding children's language environment has been of central interest to both theoretical and applied and research on early language learning.
In particular, word learning has come to be viewed as central to measuring and influencing language learning.
Important applied area, interest in sociodemographic variation

Word learning is one of the rare areas in developmental work in which the construct of individual differences is also seen as an epistemic construct: 


How is children's linguistic input related to their uptake? 
A wide range of computational models share the same basic accumulator architecture for word learning, in which hearing a word in the input leads to some increase in the probability of learning that word [@mcmurray2007; @hidaka2013; @mollica2017; @mayor2010]. 
The goal of the current research is to unify these into a "standard" model -- analogous to the "Standard Model" of physics. 
Our aim is to recapitulate the dynamics of how vocabulary grows, using words as the fundamental building blocks to understand the relationship between quantitative measures of language exposure and vocabulary size.

We take a step towards this goal by considering Bayesian models that aim to predict children's vocabulary size using inferred parameter values. 
We infer 1) the mean rate of input (tokens/hour) per child, and 2) a difficulty per word (the number of times a word must be heard to be learned). 
In our baseline model, we assume that some children hear more hourly tokens than others, that these rates are normally-distributed across children, and that the difficulty of each word is also normally-distributed. 
A second model extends the baseline model to take into account varying frequency across words (estimated from the CHILDES corpus). 
Both models assume that all children begin accumulating words at a particular *starting age*, which is to be inferred by the model. 


# The Standard Model

To build a "standard model", we begin by specifying the minimal assumptions made by any model of word learning.
First, assume that words accumulate through repeated exposure at a rate controlled by a variety of factors, including 1) learning rate, 2) frequency of exposure, 3) diversity of exposure, and 4) leveraged learning.

Combine to create a "macro-economics of early language" 
Input distributional patterns that match corpus dynamics (Montag etc)
Identify critical uptake patterns to aim for (CDI quantiles?)

@mcmurray2007 introduced a simple accumulator model of word learning to explain the vocabulary splurt... 

## Extensions
ambiguity and disambiguation
Cross-situational extensions (Smith, Blythe) - most word learning situations are ambiguous and require disambiguation across time (using mechanisms of learning and memory)
Socio-pragmatic extensions - most word learning instances are disambiguated by social and pragmatic cues (and are unambiguous)
Experimental evidence from the "associations vs. hypotheses" debate


# Procedure

## Data analysis

# Results

# Discussion


\newpage

# References
```{r create_r-references}
#r_refs(file = "references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup
